\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{tikz}
\usepackage{physics}
\usepackage{gensymb}
\usepackage{amsmath, amsfonts, mathtools, amsthm, amssymb, mathrsfs}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{hyperref}
\usepackage{float}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{import}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[a4paper, margin=1in]{geometry}
% The configuration for boxes and parts of the overall idea come from Pingbang Hu: https://github.com/sleepymalc
% Use a relative path (forward slashes or "..") to avoid backslash-escapes
\subimport{../}{header.tex}

\title{\textbf{MATH 487 Deterministic Operations Research}}
\author{GoldenOrbWeaver}
\date{Spring 2026}

\begin{document}

\maketitle

\tableofcontents{\textbf{}}

\pagebreak

\pagestyle{fancy}
\fancyhead{}
\fancyhead[L]{\thetitle}
\fancyhead[R]{GoldenOrbWeaver}

\section{Linear Programming}
\begin{definition}
    \textbf{Linear programming}: The optimization of a linear function subject to linear constraints.
\end{definition}

\begin{eg}
    Suppose a starving artist is trying to plan a food budget. He is health 
    conscious and wants a healthy diet that includes the following: at least
    70 g of protein per day, at least 100 g of carbohydrates per day, exactly
     15 mg of vitamin D per day, but no more than 75 g of fat per day.

     Five foods to choose from (fix formatting later): \\
    \begin{tabular}
    { |c | c | c | c | c | c | }
     \hline
     Food & Protein & Carbohydrates & Vitamin D & Fat & Cost \\
     \hline \\
        Hamburger & 10g/oz & 2g/oz & .5mg/oz & 8g/oz & \$0.20/oz \\
        Milk & 2g/oz & 3 g/oz & 4mg/oz & 2g/oz & \$0.02/oz \\
        Cereal & 3g/oz & 23g/oz & 2mg/oz & 1g/oz & \$0.10/oz \\
        Ch. N S & 2g/oz & 2g/oz & 0 mg/oz & 0.5g/oz & \$0.03/oz\\
        Eggs & 6g/egg & 4g/egg & 1mg/egg & 5g/egg & \$0.10/egg\\
     \hline

    \end{tabular}
     
    Question: How can he meet dietary goals while minimizing cost?
\end{eg}

\begin{answer}
Set up \textbf{decision variables}: \\
H, M, C, CNS, and E are oz (or number) per day\\

Constraints: \\
Protein: \(p=10H+2M+3C+2CNS+6E\geq70\) \\
Carbs: \( c=2H+3M+23C+2CNS+4E\geq100\) \\
Vitamin D: \(0.5H+4M+2C+E=15\) \\
Fat: \(f=8H+2M+1C+0.5CNS+5E\leq75\) \\
Nonnegativity: \(H,M,C,CNS,E\geq0\) \\

Subject to these requirements, we wish to minimize cost:
\[
    cost=20H+2M+10C+3CNS+10E
\]
\end{answer}

\begin{definition}
    Let \(f:\mathbb{R}^n\rightarrow \mathbb{R}\) be a function of n variables, then $f$ is called
    linear $\iff$ $f$ is of the form
    \[
        f(x_1,x_2, \dots x_n)=a_1x_1+a_2x_2+\dots+a_nx_n+b_0
    \]
    for some constraints
    \[
        a_1, a_2, \dots a_n \text{ and }b_0
    \]

\end{definition}

\begin{definition}
    A \textbf{linear equation} is an equation of the form \(f(x_1, \dots, x_n)=a\) constant. 
\end{definition}

\begin{definition}
    A \textbf{linear inequality} is an inequality of the form \(f(x_1, \dots, x_n)\leq\) 
    a constant, or \(f(x_1, \dots,x_n)\geq\) a constant.  
\end{definition}

\begin{definition}
    A \textbf{linear constraint} is either a linear equation or a linear inequality.
\end{definition}

\begin{definition}
    A \textbf{linear program} is the optimization of a linear function subject to linear constraints.
\end{definition}

\begin{eg}{The Furniture Problem}
    
    Suppose you are in charge of a furniture factory. Your plant makes tables
    and chairs out of iron, wood, and labor. \\
    \begin{tabular}
    { |c | c | c | c | c | }
     \hline
     Product & Iron (lbs) & Wood (ft) & Labor (hrs) & Profit (\$) \\
     \hline \\
        Table & 1 & 20 & 16 & 80 \\
     \hline \\
        Chair & 2 & 15 & 5 & 40 \\
     \hline
    \end{tabular} \\
    
    Suppose that your plant has access to 100 lbs of iron/day, 1000 lbs of 
    wood/day, and it has 80 employees and thus 640 labor hours/day. What should
    their production plan be?
\end{eg}

\begin{answer}
    First, we need to decide on the decision variables. These should have two properties:
    \begin{enumerate}
        \item The direction manager must have control over them
        \item Designation of optimal values solves the problem
    \end{enumerate}
    We select two, \(T\) and \(C\), the number of tables and chairs produced per day respectively. \\

    Next, we need to select our objective function. Since we wish to maximize
    profit, our objective function is:
    \[
        profit=\Pi =80T+40C
    \]

    We also need to figure out constraints:
    \[
        Iron: T+2C\leq100
        \\Wood: 20T+15C\leq1000 \\
        \\Labor: 16T+5C\leq640 
        \\Nonnegativity: T,C\geq0
    \]
    
    We have a linear program:
    \[
        max_{T,C} 80T+40C \\
        s.t. 
    \]

\end{answer}

\begin{remark}
    When there are two decision variables, we can graphically solve a linear program.
\end{remark}

ADD DRAWING later

\begin{definition}
    The \textbf{feasible region} of a linear program is the set of all points that
    satisfy all constraints.
\end{definition}

Geometrically, we wish to find the highest isoprofit that intersects the feasible
region, grazing the side of it. This will occur at a vertex (unless a constraint
line is parallel to the isoprofit line). We can check all the vertices or we can
analyze the slopes of the constraints and find the vertex of constraints with
slopes above and below the slope of the isoprofits. 

\begin{definition}
    An \textbf{integer program} is a linear program where all of the decision variables
    must have integer values.
\end{definition}

\begin{eg}{Blending Model}
    A scrap metal operator reviews an order for 24 lbs of tin, 15 lbs of copper, and 20 lbs of aluminum. 
    She can buy two types of scrap metal which she can melt down:
    \begin{tabular}
        { |c | c | c | c | c | }
        \hline
        Type & Tin & Copper & Aluminum & Cost (\$0.01/lb) \\
        \hline \\
        Metal 1 & 40\% & 50\% & 10\% & 20 \\
        \hline
        Metal 2 & 40\% & 10\% & 50\% & 10 \\
        \hline
    \end{tabular}
    Only 50 lbs of Metal 1 are available. How can she meet the order most effectively?
\end{eg}
\begin{answer}
    Decision variables: \\
    \(M_1\) = Amount of metal 1 to buy (lbs) \\
    \(M_2\) = Amount of metal 2 to buy (lbs)\\ 
    Linear program:\\
    min \(20M_1+10M_2=cost\)\\
    %\begin{align*}
    s.t. \(0.4M_1+0.4M_2\geq24\) (tin)\\
    \(0.5M_1+0.1M_2\geq15\) (copper)\\
    \(0.1M_1+0.5M_2\geq20\) (aluminum)\\
    \(M_1\leq50\) (availability)\\
    %\end{align*} FIX THIS
    Since there are only 2 decision variables, we can solve this graphically:
\end{answer}

\begin{eg}{Transportation Problems}

    Goods are located at sources and needed to be shipped to destinations. There 
    is a per unit cost to ship from any particular source to an particular 
    destination. The objective is to minimize the cost\dots

    Suppose that the Frank Perdue Chicken Co. has 2000 tons of chickens on inventory,
    500 of which are on a farm near San Francisco, 500 on a farm near Houston,
    and 1000 on a farm near Detroit. They wish to ship the chicken to four superstores 
    located in New York, Los Angeles, Kansas City, and Miami. Demand is NYC 300 tons,
    LA 900 tons, KC 600 tons, and Mia 200 tons. The shipping costs per ton are:\\
    \begin{tabular}
        { |c | c | c | c | c | }
        \hline
        From/To & NY & LA & KC & Mia \\
        \hline \\
        SF & 80 & 10 & 65 & 80 \\
        \hline
        Hou & 30 & 50 & 20 & 20 \\
        \hline
        Det & 30 & 100 & 50 & 50 \\
        \hline
    \end{tabular}

    Define \(x_{i,j}=\) the tonnage of chicken shipped from \(i\) to \(j\)\\
    Linear program: \\
    \[
      min_{x} 80x_{11}+10x_{12}+65x_{13}+80x_{14}+30x_{21}+50x_{12}+20x_{23}+20x_{24}+30x_{31}+100x_{32}+50x_{33}+50x_{34}     
    \]
    s.t. \(x_{11}+x_{12}+x_{13}+x_{14}\leq500\)\\
    \(x_{21}+x_{22}+x_{23}+x_{24}\leq500\)\\
    \(x_{31}+x_{32}+x_{33}+x_{34}\leq1000\)\\
    \(x_{11}+x_{21}+x_{31}\geq300\)\\
    \(x_{12+x_{22}+x_{32}\geq900}\)\\
    \(x_{13}+x_{23}+x_{33}\geq600\)\\
    \(x_{14}+x_{24}+x_{34}\geq200\)\\  
    \(x_{i,j}\geq0\) 
    \begin{remark}
        Note the specific special structure of the constraint matrix. This allows
        for specialized algorithms to solve transportation problems.
    \end{remark}
    \begin{remark}
        In this particular problem, the sum of the supplies at sources equals the 
        sum of the demands at destinations. This implies that for any feasible solution, 
        all the constraints hold with equality. In general transportation problems, 
        the total supply at sources is greater than or equal to total demand at sinks.
    \end{remark}
    \begin{remark}
        In general, the transportation problem has the form: \\
        \(max_{x}\)  \(\sum_{i=1}^{I}\sum_{j=1}^{J}C_{ij}X_{ij}\) \\
        s.t. \(\sum_{i=1}^{I}X_{ij}\leq S_i \) for \(i=1,\dots,I\) \\
        \(\sum_{j=1}^{J}X_{ij}\geq D_j \) for \(j=1,\dots,J\) \\
        \(X_{ij}\geq0\) for all \(i,j\) \\
        where \(C_{ij}=\) cost per unit shipped from source \(i\) to destination \(j\) \\ 
        \(I=\) number of sources \\
        \(J=\) number of destinations \\
        \(S_i=\) on hand at source i \\
        \(d_j=\) demand at destination j    
    \end{remark}
\end{eg}

\begin{eg}
    Daisy Drugs manufactures two drugs, Drug 1 and Drug 2. The drugs are produced
    by blending together two chemicals: Chemical ! and Chemical 2. By weight,
    Drug 1 must contain at least 65\% Chemical 1, and Drug 2 must contain at least
    55 \% Chemical 1. Both drugs, by weight, are completely composed of Chemical 1
    and Chemical 2. Drug 1 sells for \$6/oz and Drug 2 sells for \$4/oz. Chemicals  1
    and 2 are produced by one of two production processes: \\
    \begin{tabular}
        { |c | c | c | c |c | } 
        \hline
        Process & Raw Mat./hr & Labor/hr & Chemical 1 & Chemical 2 \\
        \hline \\
        I & 3 & 2 & 3 & 3 \\
        \hline 
        II & 2 & 3 & 3 & 1 \\
        \hline
    \end{tabular} \\

    A total of 120 hours of skilled labor, 100 oz of raw material are available. 
    Formulate a LP which can be used to maximize Daisy's sale revenue. 
    \begin{remark}
        There is a more complicated problem than the furniture problem because 
        we now have to model a multi-step production process. We have 1 (raw materials 
        and labor) -> 2 (chemicals) -> 3 (drugs).
    \end{remark}

    Decision variables: \\
    \(D_1=\) amount of Drug 1 produced (oz) \\
    \(D_2=\) amount of Drug 2 produced (oz) \\
    \(C_1=\) amount of Chemical 1 produced (oz) \\
    \(C_2=\) amount of Chemical 2 produced (oz) \\
    \(P_1=\) number of hours used with Process I \\
    \(P_2=\) number of hours used with Process II \\ 
    \(C_1D_1=\) amount of Chemical 1 used in Drug 1 (oz) \\
    \(C_2D_1=\) amount of Chemical 2 used in Drug 1 (oz) \\
    \(C_1D_2=\) amount of Chemical 1 used in Drug 2 (oz) \\
    \(C_2D_2=\) amount of Chemical 2 used in Drug 2 (oz) \\  

    Linear program: \\
    max \(6D_1+4D_2\) \\
    s.t. \(\frac{C_1D_1}{C_1D_1+C_2D_1}\geq0.65\) \\
    \(=>C_1D_1\geq0.65(C_1D_1+C_2D_1)\) \\
    \(=>0.35C_1D_1-0.65C_2D_1\geq0\) \\
    \(0.45C_1D_2-0.55C_2D_2\geq0\) (Since we need constraints to be linear) \\
    \(C_1-C_1D_1-C_1D_2\geq0\) \\
    \(C_2-C_2D_1-C_2D_2\geq0\) \\
    \(D_1=C_1D_1+C_2D_2\) \\
    \(D_2=C_1D_2+C_2D_2\) \\

    The constraints defining \(C_1\), \(C_1\), \(D_1\), and \(D_2\) are sometimes called 
    linking constraints. We still need constraints linking 1 to 2:

    \(C_1-3P_1-3P_2=0\) \\
    \(C_2-3P_1-P_2=0\)  \\

    We also have resource constraints: \\
    \(3P_1+2P_2\leq100\) \\
    \(2P_1+3P_2\leq120\)  \\

    And of course nonnegativity: \\
    \(D_1, D_2, C_1, C_2, C_1D_1, C_1D_2, C_2D_1, C_2D_2, P_1, P_2 \geq 0\) 

    \begin{definition}
        \textbf{The Simplex Method} is a way of solving an LP with more than 2 decision variables\dots
        
        Preliminaries: \\
        \begin{enumerate}
            \item 3 Classes of LPs 
            \begin{enumerate}
                \item LPs with finite optima
                \item Infeasible LPs, where \(F.R.=\emptyset\) 
                \item Unbounded LPs, where either:
                \begin{enumerate}
                    \item min problems where we can send \(z^*\rightarrow-\infty\) 
                    \item max problems where we can send \(z^*\rightarrow\infty\) 
                \end{enumerate}
            \end{enumerate}
            We desire that an algorithm tells us which of these classes our LP is in 
            and if an optimum exists, what it is.
        \end{enumerate}
    \end{definition}
\end{eg}

Types of linear programs: \\
\begin{enumerate}
    \item LPs in general form
    \begin{enumerate}
        \item max or min problem
        \item constraints are \(\leq\), \(\geq\), or \(=\) (never < or >) 
        \item RHS of constraints can be negative, positive or O 
        \item The variables can be nonnegative, nonpositive, or unrestricted
    \end{enumerate}
    \item LPs in standard form. Similar to LPs in general form except:
    \begin{enumerate}
        \item Constraints are all \(=\)
        \item RHS of constraints are all positive or 0
        \item Variables are all nonnegative 
    \end{enumerate}
\end{enumerate}

\begin{proposition}
    Any LP in general form can be rewritten as an equivalent LP in standard form.
\end{proposition}
\begin{proof}
    We demonstrate a procedure on an example, which can be used on any example. The 
    steps can be used in any order. 

    Put the LP below into an equivalent LP in standard form:\\
    \(max \text{ } 6x_1+x_2-x_3-x_4=z\)\\
    s.t. \(x_1-x_2\leq5\) \\
    \(x_1+2x_2+x_3-x_4\geq 2\) \\
    \(2x_2+x_3=-4\) \\
    \(x_1, x_4 \geq 0\) \\
    \(x_2\leq0\) \\
    \(x_3 \text { unrestricted}\)
    
    Step 1: Change "max" problem to "min" problem, if necessary, by multiplying the 
    objective function by -1: \\
    \(min \text{} -6x_1-x_2+x_3+x_4=-z\)
    s.t. \(x_1-x_2\leq5\) \\
    \(x_1+2x_2+x_3-x_4\geq 2\) \\
    \(2x_2+x_3=-4\) \\
    \(x_1, x_4 \geq 0\) \\
    \(x_2\leq0\) \\
    \(x_3 \text { unrestricted}\) 

    Step 2: Multiply any constraints with negative RHS by -1. (This changes 
    \(\leq\) constraints to \(\geq\) and vice versa): \\
    \(min \text{} -6x_1-x_2+x_3+x_4=-z\)
    s.t. \(x_1-x_2\leq5\) \\
    \(x_1+2x_2+2x_3-x_4\geq2\) \\
    \(-2x_2-x_3=4\) \\
    \(x_1, x_4 \geq 0\) \\
    \(x_2\leq 0\) \\
    \(x_3 \text{ unrestricted}\) \\
    
    Step 3: Add slack variables so that all constraints are \(=\): \\
    \(min \text{} -6x_1-x_2+x_3+x_4=-z\)
    s.t. \(x_1-x_2+S_1=5\) \\
    \(x_1+2x_2+2x_3-x_4-S_2=2\) \\
    \(-2x_2-x_3=4\) \\
    \(x_1, x_4\geq 0\) \\
    \(x_2\leq 0\) \\
    \(x_3 \text{ unrestricted}\) \\
    \(S_1, S_2\geq 0\) \\
    
    Step 4: Replace any nonpositive variable \(x_i\) with a new variable 
    \(x_i^*\), where \(x_i^{\prime} =-x_i\): \\
    \(min \text{} -6x_1+x_2^{\prime} +x_3+x_4=-z\)
    s.t. \(x_1+x_2^{\prime} +S_1=5\) \\
    \(x_1-2x_2^{\prime} +2x_3-x_4-S_2=2\) \\
    \(2x_2^{\prime} -x_3=4\) \\
    \(x_1, x_2^{\prime}, x_4, S_1, S_2 \geq 0\) \\
    \(x_3 \text{ unrestricted}\) \\

    Step 5: Replace any unrestricted variables \(x_i\) with the difference of two 
    nonnegative variables \(x_i^{\prime} -x_i^{\prime\prime} \): \\
    \(min \text{} -6x_1+x_2^{\prime} +x_3^{\prime} -x_3^{\prime\prime} +x_4=-z\)
    s.t. \(x_1+x_2^{\prime} +S_1=5\) \\
    \(x_1-2x_2^{\prime} +2x_3^{\prime} -2x_3^{\prime\prime} -x_4-S_2=2\) \\
    \(2x_2^{\prime} -x_3^{\prime} +x_3^{\prime\prime} =4\) \\
    \(x_1, x_2^{\prime}, x_3^{\prime} , x_3^{\prime\prime} x_4, S_1, S_2 \geq 0\) \\

    This is now in standard form

\end{proof}

\begin{note}
    Suppose we solved the linear program in standard form that we just found 
    using the simplex method and get an optimal answer of: \\
    \(x_1=0\), \(x_2^{\prime} =5\), 
    \(x_3^{\prime} =6\), \(x_3^{\prime\prime} =0\), \(x_4=0\), \(S_1=0\), \(S_2=0\), 
    \(-z=11\)\\
    Then we can translate back and find that the solution of the linear 
    program in general form is: \\
    \(x_1=0\), \(x_2=-5\), \(x_3=6\), \(x_4=0\), \(z=-11\)     
\end{note}

\begin{definition}
    An \(m\cross n\) linear program is in \textbf{canonical form} if it is 
    in standard form and there is a distinguished set of \(m\) variables called 
    \textbf{basic variables} for which: \\
    \begin{enumerate}
        \item Each basic variable has coefficient 1 in one constraint and 0 
        in the others. 
        \item Each basic variable has coefficient 0 in the objective function. 
        \item Each constraint has exactly 1 basic variable with coefficient 1.
    \end{enumerate}
\end{definition}

\begin{eg}
    Recall the furniture problem: \\
    \(\text{max }80T+40C=z\) \\
    s.t. \(T+2C\leq100\) \\
    \(20T+15C\leq1000\) \\
    \(16T+5C\leq640\) \\
    \(T,C\geq0\)\\
    
    We start by converting this to a linear program in standard form: \\
    \(\text{min }-80T-40C=-z\) \\
    s.t \(T+2C+S_1=100\) \\
    \(20T+15C+S_2=1000\) \\
    \(16T+5C+S_3=640\) \\
    \(T, C, S_1, S_2\geq0\) \\
    
    This happens to also be in canonical form with the set of basic variables 
    or basis=\(\left\{S_1, S_2, S_3\right\}\)     
\end{eg}

\begin{remark}
    We were lucky that our standard form turned out to be a canonical form. In most 
    examples this won't be the case.
\end{remark} 

\begin{remark}
    If we have a canonical form LP, it is easy to see what is called an associated \textbf{basic feasible solution}. 
    In this case, \(T=C=0\) (nonbasic variables are 0), \(S_1=100\), \(S_2=1000\), and \(S_3=640\) (basic variables are RHS).    
\end{remark}

We now have a b.f.s. for the LP, but this isn't optimal. We can raise C while reducing 
\(S_1\), \(S_2\), \(S_3\) to maintain constraints, but we can't do this indefinitely, only 
up to \(\frac{100}{2}\) for the iron constraint, \(\frac{1000}{15}\) for the wood constraint, 
and \(\frac{640}{5}\) for the labor constraint. Therefore, we should make \(C\) basic and \(S_1\) nonbasic.

\begin{theorem}{Minimum-Ratio Test}\\
    In general, if we've decided that \(x_{j^*}\) should enter the basis, then the current basic 
    variable of the \(i^*\)th row should leave the basis where \(i^*\) minimizes \(\frac{b_i}{a_{ij^*}}\) over 
    all \(i\) s.t. \(a_{ij^*}>0\) .    
\end{theorem}

To make \(C\) basic and \(S_1\) nonbasic while retaining canonical form:
\begin{enumerate}
    \item Multiply equation (1) by \(1/2\), obtaining (1').
    \item Subtract \(15*(1^{\prime})\)  from (2), obtaining (2')\\
    Subtract \(15*(1^{\prime})\) from (3), obtaining (3') \\
    Subtract \(-40*(1^{\prime})\) from (obj), obtaining (obj') 
\end{enumerate}
Doing the first step: \\
\[
    \frac{1}{2}T+C+\frac{1}{2}S_1=50
\]  
Doing the second step: \\
\begin{align}
    20T+15C+S_2&=1000 \\
    -15(\frac{1}{2}T+C+\frac{1}{2}S_1&=50 \\
    \frac{25}{2}T-\frac{15}{2}S_1+S_2&=250
\end{align}
\begin{align}
    16T+5C+S_3&=640 \\
    -5(\frac{1}{2}T+C+\frac{1}{2}S_1&=50) \\
    \frac{27}{2}T-\frac{5}{2}S_1+S_3&=390
\end{align}
\begin{align}
    -80T-40C&=-Z \\
    +40(\frac{1}{2}T+C+\frac{1}{2}S_1&=50) \\
    -60T+20S_1&=-Z+2000
\end{align}
Our new LP is: \\
\(\text{max }-60T+20S_1=-Z+2000\) \\
s.t. \(\frac{1}{2}T+C+\frac{1}{2}S_1=50\)\\
\(\frac{25}{2}T-\frac{15}{2}S_1+S_2=250\) \\
\(\frac{27}{2}T-\frac{5}{2}S_1+S_3=390\) \\
\(T, C, S_1, S_2, S_3\geq0\)\\


We have a new canonical form with basis \(C, S_2, S_3\). The associated b.f.s 
is \((T,C,S_1,S_2,S_3)=(0, 50, 0, 250, 390)\).

Note that we can read the associated \(Z\) value from the canonical form (here \(Z=2000\).) 

\begin{remark}
    In general, if we wish to enter \(x_{j*}\) into the basis and remove the basic variable 
    from the \(i^*\)th row from the basis, then the canonical form tableau changes as follows: \\
    \begin{note}
        No prime means before pivot and prime means after pivot.
    \end{note}
    \begin{enumerate}
        \item \(a_{i^*j^*}^{\prime}=1\) 
        \item \(a_{i^*j}^{\prime} =\frac{a_{i^*j}}{a_{i^*j^*}}\) for \(j=j^*\)
        \item \(a_{ij^*}^{\prime} =0\) for \(i\neq i^*\)
        \item \(a_{ij}^{\prime} =a_{ij}-\frac{a_{ij^*}*a_{i^*j}}{a_{i^*j^*}}\) for \(i\neq i^*\) and \(j\not j^*\)
        \item \(b_{i^*}^{\prime} =\frac{b_{i^*}}{a_{i^*j^*}}\)
        \item \(b_i^{\prime} =b_i-\frac{b_{i^*a_{ij^*}}}{a_{i^*j^*}}\) for \(i\neq i^*\)
        \item \(C_{j^*}^{\prime} =0\) 
        \item \(C_j^{\prime} =C_j-\frac{C_{j^*}a_{i^*j}}{a_{i^*j^*}}\) for \(j\neq j^*\)
        \item \((\pm Z)+Z_0^{\prime}=(\pm Z)+Z_0-\frac{b_{i^*}C_{j^*}}{a_{i^*j^*}}\)    
    \end{enumerate} 
    The operation which generates a new canonical form from an old canonical form is called a \textbf{pivot} on \(a_{i^*j^*}\) 
\end{remark}

An easy way to do this is to use the \textbf{box method}. First, write the original LP in 
\textbf{tableau form}: \\
\begin{tabular}{c c c c c | c}
    T & C & $S_1$ & $S_2$ & $S_3$ & RHS \\
    1 & 2 & 1 & 0 & 0 & 100 \\
\end{tabular}

\textbf{FINISH ADDING PHYSICAL NOTES}

\begin{definition}{Simplex Algorithm}

    Given an LP problem in general form:
    \begin{enumerate}
        \item Turn into an equivalent problem in standard form
        \item Obtain an initial canonical form, or else determine that the LP is infeasible. 
        \item \begin{enumerate}
            \item Consider the coefficients in the objective row of the current canonical form. If they are all nonnegative, 
            the current associate B.F.S. is optimal. If not, select a column (\(j^*\) ) in which 
            the coefficient is negative, to pivot \(x_{j^*}\) into the basis.
            \item If \(a_{ij^*}\leq0\) for \(i=i,\dots,m\), the LP is unbounded. Otherwise, 
            perform the M.R.T. to determine the coefficient \(a_{i^*j^*}\) on which to pivot next.
            \item Pivot on \(a_{i^*j^*}\), obtaining a new canonical form. Go to (a).    
        \end{enumerate}
    \end{enumerate}
\end{definition}

Suppose we are given an LP in standard form: \\
\begin{align}
    \text{min }c_1x_1+\dots+c_n x_n &=z \\
    \text{s.t. }a_{11}x_1+\dots+a_{1n}x_n&=b_1 \\
    \vdots \\
    a_{m1}x_1+\dots+a_{mn}x_n&=b_m \\
    x_{1},\dots, x_ n & \geq0
\end{align}

Let us introduce \(m\) artificial variables \(w_1,\dots,w_m\) and consider a second 
objective function as follows: \\
\begin{align}
    \text{min }c_1x_1+\dots+c_n x_n &=z \\
    \text{min }w_1+\dots+w_m &=w \\
    \text{s.t. }a_{11}x_1+\dots+a_{1n}x_n+w_1&=b_1 \\
    \vdots \\
    a_{m1}x_1+\dots+a_{mn}x_n+w_m&=b_m \\
    x_{1},\dots,x_ n, w_1, \dots, w_m & \geq0
\end{align}

We note two things: \\
\begin{enumerate}
    \item The original LP is feasible iff the optimal \(w^*\) is 0.
    \item The LP (where \(w\) is the objective) is almost in canonical form. The only 
    problem is that we have \(w_i\)s in the objective function. 
\end{enumerate}

To fix (2), we use the substitutions:
\begin{align}
    w_1=b_1-a_{11}x_1-a_{12}x_2-\dots-a_{1n}x_n \\
    \vdots \\
    w_m=b_m-a_{1m}x_1-a_{m2}x_2-\dots-a_{mn}x_n \\
\end{align}

So the second objective function is now: \\
\begin{align}
    \text{min }w&=w_1+\dots+w_m \\
    &= \sum_{i=1}^m b_1-(\sum_{i=1}^m a_{11})x_1-\dots -(\sum_{i=1}^m a_{1n})x_n
\end{align}

We rewrite this: \\
\[
    \text{min }-(\sum_{i=1}^m a_{11})x_1-\dots -(\sum_{i=1}^m a_{1n})x_n=w-\sum_{i=1}^m b_1
\]

So, part 2 of the simplex method is: Given the LP in standard for, set up the following 
LP: \\
\begin{align}
    \text{min }c_1x_1+\dots+c_n x_n &=z \\
    \text{min }-(\sum_{i=1}^m a_{11})x_1-\dots -(\sum_{i=1}^m a_{1n})x_n &=w-\sum_{i=1}^m b_1 \\
    \text{s.t. }a_{11}x_1+\dots+a_{1n}x_n+w_1&=b_1 \\
    \vdots \\
    a_{m1}x_1+\dots+a_{mn}x_n+w_m&=b_m \\
    x_{1},\dots,x_ n, w_1, \dots, w_m & \geq0
\end{align}

This is in canonical form with basis \(w_1,\dots,w_m\)

We solve this to find the optimal \(w^*\). We have two scenarios: \\
\begin{enumerate}
    \item If \(w^*>0\), the LP is infeasible \\
    \item If \(w^*=0\), the LP is feasible and you have a canonical form 
\end{enumerate} 

\begin{eg}
    \begin{align}
        \text{min }3x_1+x_2-x_3-x_4 &=z \\
        \text{s.t. }x_1-x_2+x_3+2x_4 &=3 \\
        2x_1+2x_2-x_3-x_4 &=1 \\
        -x_1+x_2+3x_3+3x_4 &=6 \\
        x_1, x_2, x_3, x_4 &\geq 0
    \end{align}
\end{eg}
\begin{answer}
    First, we add \(m=3\) artificial variables \(w_1, w_2, w_3\): \\
    \begin{align}
        \text{min }3x_1+x_2-x_3-x_4 &=z \\
        \text{min }-2x_1-2x_2-3x_3+x_4 &=w-10 \\
        \text{s.t. }x_1-x_2+x_3+2x_4 +w_1 &=3 \\
        2x_1+2x_2-x_3-x_4 +w_2&=1 \\
        -x_1+x_2+3x_3+3x_4 +w_3&=6 \\
        x_1, x_2, x_3, x_4, w_1, w_2, w_3 &\geq 0
    \end{align}

    We write this in tableau form: \\
    \begin{tabular}{c c c c c c c | c }
        \(x_1\) & \(x_2\) & \(x_3\) & \(x_4\) & \(w_1\) & \(w_2\) & \(w_3\) & RHS \\
        1 & -1 & 1 & -2 & 1 & 0 & 0 & 3 \\
        2 & 2 & -1 & -2 & 0 & 1 & 0 & 1 \\
        -1 & 1 & 3 & 3 & 0 & 0 & 1 & 6 \\
        \hline
        -2 & -2 & -3 & 1 & 0 & 0 & 0 & \(w-10\) \\
        3 & 1 & -1 & -1 & 0 & 0 & 0 & z
    \end{tabular}

    We perform the M.R.T., \(\frac{3}{1}\) vs \(\frac{1}{2}\), and chose \(\frac{1}{2}\). Thus, we pivot on \(x_1\): \\
    
    %\begin{tabular}
       % \(x_1\) & \(x_2\) & \(x_3\) & \(x_4\) & \(w_1\) & \(w_2\) & \(w_3\) & RHS \\
      %  0 & -2 & $\frac{3}{2}$ & -1 & 1 & $\frac{-1}{2}$ & 0 & $\frac{5}{2}$ \\
     %   2 & 2 & -1 & -2 & 0 & 1 & 0 & 1 \\
    %    -1 & 1 & 3 & 3 & 0 & 0 & 1 & 6 \\
   %     \hline
  %      -2 & -2 & -3 & 1 & 0 & 0 & 0 & \(w-10\) \\
 %       3 & 1 & -1 & -1 & 0 & 0 & 0 & z
 %   \end{tabular}

    \textbf{finish writing tableau for pivots}


    Since \(w=0\), the LP is feasible. However, \(z\) is not optimized because we 
    still have a negative coefficient in the objective function. Therefore, we pivot on (FINSIH THIS)
\end{answer}

\begin{definition}{\textbf{Multi-period Decision Models}}

    These are dynamic problems in where decisions must be 
    made over a series of time periods. Many of these problems can also be solved using \textbf{dynamic programming}\dots

    Idea of formulation: To formulate such problems as LPs, you must include linking constraints, which tie 
    together variables from one period to variables from previous periods.
\end{definition}

\begin{eg}{Staffing a Training Program}



\end{eg}
\begin{answer}
    We first chose our decision variables: \\
    \begin{align*}
        A_t&= \text{number of agents available at the beginning of each month t} \\
        H_t&= \text{number of students hired at beginning of month t}
    \end{align*}

    Then, we formulate an LP: \\
    \begin{align*}
        \text{min }350(H_1+\dots+h_6)+600(A_1+\dots+A_6) \\
        \text{s.t. }A_1&=60 \\
        150A_1-20H_1&\geq8000 \\
        150A_2-20H_2&\geq 9000 \\
        \vdots \\
        150A_6-20H_6&\geq11000 \\
    \end{align*}
\end{answer}
\textbf{FINISH WRITING THIS OUT}

\begin{eg}
    Suppose a salt mining company in Utah makes salt for use in clearing
    snowy streets. It has 20 tons of salt on hand as of July 1st. It forecasts
    its future as follows: \\
    \begin{tabular}{| c | c | c |}
        \hline
        Period & Amount of salt & Costs (\$/ton) \\
        \hline
        Jul-Sep & 10 & 100 \\
        \hline
        Oct-Dec & 40 & 300 \\
        \hline
        Jan-March & 80 & 500 \\
        \hline
        Apr-Jun & 40 & 250 \\
        \hline
    \end{tabular}

    In addition to meeting its demand, the company must also balance the books 
    by having exactly 20 tons on hand the following July 1st.

    It costs \$50 to store a ton of salt on company grounds through the end of
    a period. Formulate an LP to minimize costs.
\end{eg}
\begin{answer}
    
    
    In problems like this, it is usually beneficial to: \\
    \begin{enumerate}
        \item Index the decision periods
        \item Have a diagram of a decision period indicating what gets done
        when. This can either be given or assumed.
    \end{enumerate}
    
    We start by choosing the decision variables: \\
    \begin{align*}
        S_i&=\text{tons of salt produced in period i}\in(1,2,3,4) \\
        I_i&=\text{amount of inventory after demand + production at end of period i}\in(0,1,2,3,4)
    \end{align*}

    Then, we formulate the LP: \\
    \begin{align*}
        \text{min }100S_1+300S_2+500S_3+250S_4+50(I_1+I_2+I_3+I_4)&=z\\
        \text{s.t. }I_0&=20 \\
        I_1=I_0+S_1-10 \\
        I_2=I_1+S_2-40 \\
        I_3=I_2+S_3-80 \\
        I_4=I_3+S_4-40\\
        I_4=20 \\
        S_1,S_2,S_3,S-4,I_0,I_1,I_2,I_3,I_4\geq0
    \end{align*}

    \begin{note}
        In practice the salt maker might:
        \begin{enumerate}
            \item Solve the LP paying strict attention to \(I_1\) and \(S_1\)
            \item Do their July-September business according to \(I_1\) and \(S_1\)
            \item In September, make an updated forecast of demand for October to December, 
            January to March, April to June, and July to September.
            \item Reformulate a new LP according to the new demand. 
            \item Solve the new LP, obtaining an October to December strategy, etc. 
        \end{enumerate} 

        This is called a rolling planning horizon.
    \end{note}
\end{answer}

\begin{eg}
    Suppose backlogging is allowed (not strictly required to meet demand), 
    but a penalty cost of \$400/ton short is assessed. The shortage is then carried 
    forward to the next period.
    
    Another possibility is that the shortage is wiped off the slate.
\end{eg}
\begin{answer}
    We first choose our decision variables: \\
    \begin{align*}
        S_i&=\text{tons of salt produced in period i} \\
        I_i^{\prime} &=\text{amount of positive inventory at end of period i after demand} \\
        I_i^{\prime\prime} &=\text{ amount of negative inventory at end of period i, after demand}
    \end{align*}

    \begin{note}
        In terms of the previous LP: \\
        \[
            I_i=I_i^{\prime} -I_i^{\prime\prime} 
        \]

        If we set the penalty cost to \(\infty \)\$/ton, we should get the same optimum 
        solution as the previous problem. 
    \end{note}

    We formulate the LP: \\
    \begin{align*}
        \text{min }100S_1+300S_2+500S_3+250S_4+50\sum_{i=1}^4I_i^{\prime} +400\sum_{i=1}^4I_i^{\prime\prime} \\
        \text{s.t. }I_0^{\prime}-I_0^{\prime\prime}&=20 \\
        I_1^{\prime} -I_1^{\prime\prime} &=I_0^{\prime} +I_0^{\prime\prime} +S_1-10 \\
        I_2^{\prime} -I_2^{\prime\prime} &=I_1^{\prime} -I_1^{\prime\prime} +S_2-40 \\
        I_3^{\prime} -I_3^{\prime\prime} &=I_2^{\prime} -I_2^{\prime\prime} +S_3-80 \\
        I_4^{\prime} -I_4^{\prime\prime} &=I_3^{\prime} -I_3^{\prime\prime} +S_4-40 \\
        I_4^{\prime}&=20 \\
        I_4^{\prime\prime}&=0 \\
        \text{All variables nonnegative}
    \end{align*}

    \begin{note}
        It would be impossible for \(I_i^{\prime} \) and \(I_i^{\prime\prime} \), 
        for any particular \(i\), to both be positive in an optimal solution.   
    \end{note}
\end{answer}

Sensitivity Analysis: \\
We go back to the furniture problem
\begin{align*}
    \text{max }80T+40C&=z \\
    \text{s.t. }T+2C&\leq100 \\
    20T+15C&\leq1000 \\
    16T+5C&\leq640 \\
    T,C&\geq0 \\
\end{align*}

Question: Suppose \(x\) more labor hours become available. In terms of \(x\), 
by how much (if any) will this raise our \(z^*\)?

Graphically, this is equivalent to shifting the labor constraint outward. If 
\(x\) is relatively small, the maximum will still be where the labor constrain and 
the wood constraint intersect, so the optimum would still be at \((T,C)\) where 
\(20T+15C=1000\) and \(16T+5C=640+x\).

We solve this and find that: \\
\begin{align*}
    T&=\frac{230}{7}+\frac{3}{28}x \\
    C&=\frac{160}{7}-\frac{1}{7}x
\end{align*}

Therefore, the new optimum is \((32 \frac{6}{7}+\frac{3}{28}x, 22 \frac{6}{7}-\frac{1}{7}x)\), 
which agrees with our graphical picture. 

The new optimal objective function \(z^*\) is equal to \(3542 \frac{6}{7}+\frac{20}{7}x\)\dots

Conclusion: If \(x\) is the increase in labor capacity, we gain \(\frac{20}{7}x\) in optimal profits. 

\begin{definition}
    The \textbf{dual price} or \textbf{shadow price} of a constraint is the amount of improvement in the 
    optimal objective function value given a unit increase in the RHS of that constraint.
\end{definition}

A constraint with a nonzero "s lack or surplus" has a dual price of 0.

\begin{eg}
    WRITE NEW HAVEN BUSING BACKGROUND
\end{eg}
\begin{answer}
    To formulate a linear program, we start by choosing our decision variables: \\
    
    \begin{align*}
        w_{ij}&=\text{\# of white students bused from district i to school j} \\
        m_{ij}&=\text{\# of minority students bused from district i to school j} \\
    \end{align*}

    We then choose our objective function, the number of bus-miles driven: \\
    \[
        z(w_{ij}, m_{ij})=\sum_{i=1}^{30} \sum_{j=1}^{4}c_{ij}*(w_{ij}+m_{ij}) \text{ where \(c_{ij}\) is the distance from i to j}
    \]

    We then formulate the constraints. First, everyone must go to school, so: \\
    \begin{align*}
        \sum_{j=1}^4 w_{ij}&=W_i \text{ for }i=1,\dots,30 \\
        \sum_{j=1}^4 m_{ij}&=M_i \text{ for }i=1,\dots,30 \\
    \end{align*}

    We also need to not exceed the capacities of the schools: \\
    \begin{align*}
        tw_j+tm_j&\leq C_j \text{ for }j=1,\dots,4
    \end{align*}
    where \\
    \begin{align*}
        tw_j&=\sum_{i=1}^{30} w_{ij} \\
        tm_j&=\sum_{i=1}^{30} m_{ij}
    \end{align*}

    Additionally, the \% of minority students at each school must be between 
    29.05\% and 53.95\%: \\
    \begin{align*}
        0.2905 &\leq \frac{tm_j}{tw_j+tm_j} \leq 0.5395 \text{ for }j=1,\dots,4 \\
        \Rightarrow 0.2905(tw_j+tm_j) &\leq tm_j \leq 0.5395(tw_j+tm_j) \text{ for }j=1,\dots,4 \\
        \Rightarrow -0.7095tm_j+0.2905tw_j &\leq 0 \text{ for }j=1,\dots,4 \\
        \text{and } 0.4605tm_j-0.5395tw_j &\leq 0 \text{ for }j=1,\dots,4   
    \end{align*}

    And, of course, nonnegativity.
\end{answer}

Q: Can an equality constraint have a shadow price?

A: Yes, but you need to be careful to interpret it using the precise definition given in class. 

\begin{eg}
    Vitamin Problem: \\
    \begin{align*}
        \text{min }20H+2M+10C+3CNS+10E&=\text{cost} \\
        \text{s.t. }10H+3M+3C+2CNS+6E&\geq 70 \text{(protein)}\\
        2H+3M+23C+2CNS+4E&\geq 100 \text{carbohydrates)} \\
        0.5H+4M+2C+E&=15 \text{(vitamin D)} \\
        8H+2M+C+0.5CNS+5E&\leq 75 \text{(fat)} \\
        H,M,C,CNS,E&\geq 0
    \end{align*}
\end{eg}
\begin{answer}
    Raising the 15 on the RHS of row 4 to \(15+x\) (for \(-12\leq x\leq105\)) will decrease 
    the optimal objective value by 0.326923 
\end{answer}

Q: Can a shadow price be negative? 

A: Yes, it just means that increasing the RHS of a constraint yields a disimprovement in the 
objective function value, which typically happens with a \(\geq\) constraint, but can also happen 
with an \(=\). This occurs with the protein and carbohydrate constraints in the diet problem. 

\subsection{Sensitivity of objective function coefficients}
\begin{align*}
    Furniture problem: \\
    \text{max }80T+40C \\
    \text{s.t. }T+2C&\leq 100 \\
    20T+15C &\leq 1000 \\
    16T+5C &\leq 690 \\
    T,C &\geq 0
\end{align*}

Q: Suppose the profit per table changes from \$80 to \$\(80+x\). How does that affect \(z^*\)? 

A: Suppose \(x\) is very close to 0. Then the slope of the isoprofit lines will be 
\(\frac{-(80+x)}{40}\). As long as this stays between \(\frac{-4}{3}\) and \(\frac{-16}{5}\), the point 
\((32 \frac{6}{7}, 22 \frac{6}{7})\) will remain optimal. However, the optimal \(z^*\) will increase by 
\(32 \frac{6}{7}*x\).

Q: Over what range of \(x\) values is the above true?

A: So long as \(\frac{-16}{5} \leq \frac{-(80+x)}{40} \leq \frac{-4}{3}\Rightarrow 48 \geq x \geq -26 \frac{2}{3}\). These values 
can be found on the LINDO printout as the current allowable increases and decreases.

\subsection{Pricing Out a New Variable}

\textbf{ASK FOR NOTES FROM WEDNESDAY 2/25}

Suppose the furniture company now considers making bookshelves. Each bookshelf requires 
4 pounds of iron, 25 pounds of wood, and 8 hours of labor. They can bring in \$100 profit per unit. Should 
the company produce bookshelves.

Analysis: \\
Suppose the company produces 1 bookshelf but otherwise produces all tables and chairs. This would 
lower the RHS of the original furniture problem by 4, 25, and 8, respectively. From the LINDO printout, 
lowering the RHS of the iron constraint by 4 causes the \(z^*\) to decrease by \(4*0=0\). Lowering the RHS 
of the wood constraint causes \(z^*\) to drop by \(25*1.714=42.86\). Lowering the RHS of the labor constraint by 
8 causes \(z^*\) to drop by \(8*2.857=22.86\). Thus, the total loss is \(0+42.86+22.86=65.72\). Since this is less 
than the profit of \$100, the company should produce bookshelves. 

This procedure is called \textbf{pricing out} a new variable.

\begin{eg}
    Suppose that the starving artist in the diet problem considers buying bread. Bread contains 
    no protein, 20 g of carbohydrates, 1 mg of vitamin D, and 4 g of fat per oz. How much is he willing 
    to pay per oz of bread?
\end{eg}
\begin{answer}
    Consuming one oz of bread lowers the RHS of the diet problem by 0, 20, 1, and 4 respectively. 
    This disimproves the optimal objective function value by \(0*(-1.192)+20*(-0.308)+1*(0.327)+4*0=-5.833\). This means it improves the 
    optimal objective function value by +5.833, meaning the cost lowers by 5.833 cents. Therefore, the artist should be willing to pay 
    up to 5.833 cents per oz of bread. 
\end{answer}

The optimal solution includes no eggs. How much would it cost the starving artist to 
eat eggs?

We can solve this by pricing out eggs.

\begin{definition}
    For any variable \(y\) already in an LP, the reduced cost of the variable is 
    the disimprovement in the optimal objective function value per unit of \(y\) required in 
    the solution.   
\end{definition}

\begin{remark}
    If \(y\) is already a basic variable in an optimal solution, the reduced cost is 0. 
\end{remark}

\section{Duality Theory}
\textbf{ADD NOTES FROM 3/4}
Rules for generating dual linear program, given an LP: \\
Max problem | Min problem (reformat in columns):\\
\(i^{th}\) constraint \(\leq \) | \(i^{th}\) variable \(\geq 0\) \\
\(i^{th}\) constraint \(=\) | \(i^{th}\) variable unrestricted  \\
i^{th} constraint \(\geq\) | i^{th} variable \(\leq 0\) \\
j^{th} variable \(\geq 0\) | j^{th} constraint \(\geq \) \\
j^{th} variable unrestricted | j^{th} constraint \(=\)    \\
j^{th} variable \(\leq 0\) | j^{th} constraint \(\leq \) \\
Objective function coefficients | RHSs\\
RHSs | Objective function coefficients \\
Coefficients matrix A | coefficient matrix \(A^T\) \\

\begin{eg}
    Given: \\
    \begin{align*}
        \text{min }2x_1+3x_2-x_3-x_4&=z \\
        \text{s.t. }x_1+x_2+x_3 &=5 \\
        x_1-2x_2+4x_4&\geq 6 \\
        x_1, x_4 &\geq 0 \\
        x_2 \text{ unrestricted} \\
        x_3&\leq 0
    \end{align*}
    Find the dual
\end{eg}
\begin{answer}
    \begin{align*}
        \text{max }5y_1+6y_2 \\
        \text{s.t. }y_1+y_2 &\leq 2 \\
        y_1-2y_2 &=3 \\
        y_1&\geq -1 \\
        4y_2&\leq -1 \\
        y_1 \text{ unrestricted} \\
        y_1&\geq 0
    \end{align*} 
\end{answer}

What happens if we take the dual of the dual?
\begin{align*}
    \text{min }2z_1+3z_2-z_3-z_4 \\
    \text{s.t. }z_1+z_2+z_3 &= 5 \\
    z_1-2z_2 +4z_4 &\geq 6 \\
    z_1,z_4&\geq 0 \\
    z_2 \text{ unrestricted} \\
    z_3&\leq 0
\end{align*}

If an LP (B) is the dual of an LP (A), then LP(A) is the dual of LP(B).

\begin{theorem}{\textbf{Weak Duality Theorem}}
    The objective value of the dual LP at any feasible solution is always a bound 
    on the objective of the primal LP at any feasible solution (upper or 
    lower bound, depending on whether it is a maximization or minimization problem). 
    In fact, this bounding property holds for the optimal values of the dual and primal 
    LPs.
\end{theorem}

\begin{theorem}{\textbf{Strong Duality Theorem}}
    If the primal has an optimal solution then the dual has an optimal solution too,
     and the two optima are equal.
\end{theorem}

The optimal dual variables are the shadow prices for the primal LP.

The allowable increases and decreases for the constraints of one program are those for the objective 
function coefficients in the other.

\begin{theorem}
    Suppose (LP) is a linear program with dual (D). Then, one of the following four conditions must hold: \\
    \begin{enumerate}
        \item (LP) and (D) are both infeasible. 
        \item (LP) is unbounded, which implies that (D) is infeasible. 
        \item (D) is unbounded, which implies that (LP) is infeasible. 
        \item Both (LP) and (D) have optimal solutions with finite optimal objective function values \(z^*\) and \(z^*\). 
        In this case, \(z^*=z^*\).   
    \end{enumerate}
\end{theorem}

\begin{theorem}{\textbf{Complimentary Slackness Theorem}}
    Suppose we have the following LP and its dual: \\
    \begin{align*}
        \text{max } \vec{c}^T \vec{x} \\
        \text{s.t. }A\vec{x}&\leq \vec{b} \\
        \vec{x}&\geq \vec{0}
    \end{align*}
    Dual: \\
    \begin{align*}
        \text{min }\vec{b}^T\vec{y} \\
        \text{s.t. }A^T\vec{y}^&\geq \vec{c} \\
        \vec{y}&\geq 0
    \end{align*}
    Then: \\
    \begin{align*}
        \vec{x}^* \text{ is optimal for (P)} \\
        \vec{y}^* \text{ is optimal for (D)}
    \end{align*}
    if and only if: \\
    \begin{align*}
        x^*_j(A^T\vec{y}^*-\vec{c})_j=0 \text{ for }j=1,2,\dots \\
        y^*_i(A\vec{x}^*-\vec{b})_i=0 \text{ for } i=1,2, \dots
    \end{align*}
\end{theorem}

\begin{section}{Integer Programming}
Now we consider the case where one or more of the decision variables is constrained to be an integer. 

\begin{eg}
    \begin{align*}
        \text{max }x_1+3x_2&=z \\
        \text{s.t. }2x_1+x_2&\leq 6 \\
        x_1+5x_2&\leq 10 \\
        x_1,x_2&\geq 0
    \end{align*}
\end{eg}
\begin{answer}
    \textbf{Add this later. Also fix formatting for -th}
\end{answer}

(ask about where i should look since i can't take 751)

\begin{eg}{\textbf{Altec Case}}
    Altec inc. is considering two options for expansion to accommodate up to four new machines: 
    the vacuum melting furnace, the ingot press, the radiac trimmer, and the extrusion press.
\end{eg}
\begin{answer}
    We start by choosing our decision variables: \\
    \begin{align*}
        e&=\begin{cases}
            1 & \text{ if you build extension} \\ 
            0 & \text{ if not} 
        \end{cases} \\
        a&=\begin{cases}
            1 & \text{ if you build alcove} \\
            0 & \text{ if not}
        \end{cases} \\
        x_1&=\begin{cases}
            1 & \text{ if VMF purchased} \\
            0 & \text{ if not}
        \end{cases}\\
        x_2&=\begin{cases}
            1 & \text{ if IP purchased} \\
            0 & \text{ if not}
        \end{cases}\\
        x_3&=\begin{cases}
            1 & \text{ if RT purchased} \\
            0 & \text{ if not}
        \end{cases}\\
        x_4&=\begin{cases}
            1 & \text{ if LEP purchased} \\
            0 & \text{ if not}
        \end{cases}\\
    \end{align*}

    Since we want to maximize the net present value profits, our objective function is 
    \[
        \text{max }22x_1+52x_2+40x_3+61x_4-10a-30e
    \]

    We also need constraints for space and power: 
    \begin{align*}
        600x_1+1400x_2+800x_3+2000x_4 &\leq 400a+1000e+1200 \\
        5x_1+10x_2+10x_3+15x_4\leq 5a+20e
    \end{align*}

    And our variables must equal 0 or 1: \\
    \[
        x_1,x_2,x_3,x_4,a,e =0 \text{ or } 1
    \]

    It turns out that the solution to this LP is \(x_2^*=x_3^*=e^*=1, x_1^*=x_4^*=a^*=0\) since this 
    fits to the constraints the tightest. 
\end{answer}

\section{Networks and Graph Theory}
\begin{definition}
    An undirected graph \(G=(V, E)\) consists of a set of vertices \(V\) and a set of 
    edges \(E\) connecting pairs of the vertices in \(V\).  
\end{definition}

\begin{eg}
    \begin{align*}
        V&={A,B,C,D,E} \\
        E&={AB, CD, AD, BC}
    \end{align*}

    We represent the graph \(G=(V,E)\) with a diagram: 

    \begin{tikzpicture}[shorten >=1pt,->]
        \tikzstyle{vertex}=[circle,fill=black!25,minimum size=12pt,inner sep=2pt]
        \node[vertex] (A) at (-1,-1) {A};
        \node[vertex] (B) at (0,0)   {B};
        \node[vertex] (C) at (1,-1)  {C};
        \node[vertex] (D) at (2, -1) {D};
        \node[vertex] (E) at (1, -2) {E};
        \draw (A) -- (B) -- (C) -- (D) -- cycle;
    \end{tikzpicture}
\end{eg}

An edge where both vertices are the same is called a \textbf{loop}. 

\begin{definition}
    A \textbf{directed graph} or \textbf{digraph} is a system \(D=(V,A)\) where \(V\) is 
    a set of vertices and \(A\) is a set of \textbf{directed arcs} connecting pairs of vertices in \(V\). 
\end{definition}

\begin{eg}
    \begin{align*}
        V&={A,B,C,D,E,F} \\
        A&={\vec{AB}, \vec{BB}, \vec{CD}, \vec{BF}, vec{DC}}
    \end{align*}

    \textbf{Draw this in TekZ!!!!!!}
\end{eg}

Shortest path problems: \\
Given a digraph \(D=(v,a)\) with nonnegative weights \(w_{ij}\) on each \(\vec{y}\in a\) and 
designated vertices \(s,t\in V\), find the shortest path from \(s\) to \(t\).  

Another application:\\
\begin{eg}
    We want to drive a Buick Century for the next 5 years. The following is true of the model: \\
    \begin{tabular}{c | c | c }
        Age & Max Cost for Next Yr (1000\$) & Trade-in Value (1000\$)\\
        \hline
        0 & 2 & 12 \\
        1 & 4 & 7 \\
        2 & 5 & 6 \\
        3 & 9 & 2 \\
        4 & 12 & 1 \\
        5 & 12 & 0 \\
    \end{tabular}

    We want to minimize the total costs of driving a Buick Century over the next 5 years (from 
    beginning of year 0 to beginning of year 5). There are many options for when you replace your car.
\end{eg}
\begin{answer}
    We notice that we can represent potential plans of action as paths in a digraph. We have nodes for 
    each year and multiple arcs between each, with each arc representing a different option (such as keeping 
    the car for one year, keeping it for two years, etc.).

    \textbf{DRAW THIS DIAGRAM}

    Then, we can run Dijkstra's algorithm on this to find the shortest path. The result is 
    0 to 2 to 4 to 5, with a length of 31 (or cost of \$31,000).
\end{answer}

\begin{definition}{\textbf{Dijkstra's Algorithm}} 
    Given digraph \(D=(v,a)\), vertices \(s, t\in v\): \\
    \begin{enumerate}
        \item Set \(P={s}\), \(p=s\), \(D=(d_s, d_1, d_2, \dots, d_n, d_t)=(0, \infty, \infty, \dots, \infty)\), 
            \(A=(x-, a_1, a_2, \dots, a_n, a_t)=(x, \emptyset, \emptyset, \emptyset, \emptyset)\) 
        \item Given \(p\), consider each arc \(\vec{pq}\in a\) for which \(q \notin P\) 
            \begin{enumerate}
                \item If \(d_q>d_p+\abs{\vec{pq}}\), change \(d_q\) to \(d_p+\abs{\vec{pq}}\) and set \(a_q=p\). 
                \item If \(d_q<d_p+\abs{\vec{pq}}\), do nothing.     
            \end{enumerate}
        \item Let \(p\) be an element \(\notin P\) for which \(d_p\) is smallest. 
            \begin{enumerate}
                \item If \(p=t\), we found the shortest path.
                \item If \(p\neq t\), set \(P=P \cup {p}\) and go to 2.   
            \end{enumerate}   
    \end{enumerate}

    \textbf{FIX FORMATTING HERE AND EARLIER SO COMPILATION WORKS}
\end{definition}

\begin{eg}

\end{eg}

Suppose D=(v,a)

\textbf{EOQ Model:}\\
\begin{align*}
    \text{Ordering Cost: } C(Q)&=K+cQ \\
    \text{Holding Cost: } H(z)&=hz \text{ per time unit} \\
    \text{Demand Rate: } &=a \text{ per unit time}\\
    \text{Penalty Cost: } \infty \text{ if shortages not permitted}, P(z)=Pz \text{ per unit time}\\
\end{align*}

Diagram: \\

Case where shortages are not permitted is a special case of more general case, just with the 
penalty cost set to \(\infty \). When we take the limit of \(Q*\) for \(p\rightarrow \infty \), we should get the 
formula for the formula for \(Q^*\) with shortages not permitted.   

Also, we had an example where a steel company with \(K=\$5000\), \(c=30\), \(p=\$50\), \(h=\$20\), 
and \(a=500\). We got \(Q^*=591.6\), \(\frac{Q^*}{a}=1.18\), \(S^*=422.6\), and \(S^*-Q^*\approx 169.0\)\dots

Two generalizations of EOQ model: \\
\begin{enumerate}
    \item Quantity discounts: Suppose the cost is given by \(C(z)=\begin{cases} k+c_1z \text{ if } z<b_1 \\ \dots \\ k+c_{n+1}z \text{ if } z\geq b_n \end{cases}\) where 
    \(b_1<b_2<\dots<b_n\) and \(c_1>c_2>\dots>c_{n+1}\) 
    \item Known fixed time delivery lag: If delivery is not instantaneous and instead doesn't occur 
    until \(\lambda \) time units have passed, then we compensate by just ordering \(\lambda\) time units sooner, or at time \(t^*-\lambda \) 
    instead of  \(t^*\). This turns out to be when the inventory is \(S^*-Q^*+a \lambda \mod{Q^*/a}\)  
\end{enumerate}

Thus, if in the steel company example, there is a 3 month delay, then we should order when the inventory level is 
\(422.5-591.6+500*0.64=151.0\).

\section{Dynamic Programming (DP)}
DP is another solution technique for solving certain types of problems which occur over stages. Many times 
an important feature is starting at the end and working backwards.

\begin{eg}
    A Navy recruiter has 5 days to recruit. She starts in New York and will then go to Chicago and 
    Los Angeles. The number of potential recruits she can contact in each city is a function of the 
    number of (whole) days she spends there: 
    \begin{tabular}{c | c | c | c }
        Days & New York & Chicago & Los Angeles \\
        \hline
        0 & 0 & 0 & 0 \\
        1 & 45 & 20 & 50 \\
        2 & 70 & 45 & 70 \\
        3 & 90 & 75 & 80 \\
        4 & 105 & 110 & 100 \\
        5 & 120 & 150 & 130 \\
    \end{tabular}

    How much time should she spend in each city to maximize the total number of potential recruits contacted?
\end{eg}
\begin{answer}
    This problem has three stages: NY (\(i=1\)), Chi (\(i=2\)), and LA (\(i=3\)). We start by setting up the variables: \\
    \[
        V_i(x)=\text{max number of recruits contacted given recruiter goes to city i with x days left}
    \]
    At each stage, there are 6 possible states (for \(x\) from \(0\) to \(5\)). We are trying to calculate: \(V_1(5)\) 

    Observe that the definition of \(V_i(x)\) obeys the following recursion: \\
    \[
        V_i(x)=\max{x_i=0,1,\dots, x, [V_i(x_i)+V_{i+1}(x-x_i)]} %REFORMAT THIS MAX SHOULD BE WHOLE THING
    \]  

    We can use this formula to calculate \(V_2(x)\) from \(V_3(x)\) and then calculate \(V_1(x)\) from \(V_2(x)\). Thus, 
    we can just work backwards. 

    Doing this, we find that the optimal strategy (highest \(V_1(5)\)) is to spend 1 days in NY, 3 days in Chicago, and 1 day in LA, for a total of 170 potential recruits contacted.
\end{answer}

 \textbf{Basic setup for a finite-stage, discrete-state, deterministic dynamic programming problem:} \\
 \begin{enumerate}
    \item The problem can be divided into stages, with a decision required at each stage. In many problems, 
    the stages correspond to different time periods.
    \item Each stage can have one or more states associated with it. 
    \begin{enumerate}
        The problem can usually be easily stated in terms of finding the value or worth \(V_i(x)\) of a particular 
    \end{enumerate}
 \end{enumerate}

 ADD STUFF FROM PAPER NOTES


\end{document}